Parsing Postfix log files

### 2009-04-06 doc/logparser-complications.tex
Changes from proofreading of introduction chapter.
Added a new action, DELIVERY_ERROR, to save data from delivery errors that     happen before a disconnection. Changed some UNINTERESTING rules to use DELIVERY_ERROR and SAVE_DATA. Removed connection_data fields from UNINTERESTING rules. Added a debug field to rules, so that individual rules can have debugging     information dumped when they recognise a log line. Improved the abstract, though it still needs work. Documented everything in the thesis.
Added a table showing mean and stddev for the number of invocations of each action, for differe

\section{Complications Encountered While Writing \parsernamelong{}}

@@

@@

    \item Most importantly: every section of code must know which state
        table it should query.

@@

Once again Postfix does not log a message making the client's behaviour
clear, so once again heuristics are required to identify when this
behaviour occurs.  A list of all mails accepted during a connection is
saved in the connection's state table entry, and the accepted mails are
examined when the disconnection action is invoked.  Each mail is checked
for the following:

@@

Some mail clients disconnect abruptly if a second or subsequent recipient
is rejected; they may also disconnect after other errors, but such
disconnections are either unimportant or are handled elsewhere in the
parser, e.g.\ \sectionref{timeouts during data phase}.  Sadly, Postfix does
not log a message saying the mail has been discarded, as should be expected
by now.  The checks to identify this happening are:

@@

The author has only observed this complication occurring after a timeout,
though there may be other circumstances that trigger it.  Sometimes the
\daemon{cleanup} log line for a mail being accepted is logged after the
timeout log line, by which time \parsername{} has discarded the mail;
parsing the \daemon{cleanup} log line causes the
\action{CLEANUP\_PROCESSING} action to create a new state table entry, to
help deal with re-injected mails (\sectionref{Re-injected mails}).  This is
incorrect because the log line actually belongs to the mail that has just
been discarded; if the queueid is reused there will be a queueid clash,
otherwise the new mail will just remain in the state tables.

@@

\subsection{Out of Order Log Lines}

@@

\subsection{Mail Deleted Before Delivery is Attempted}

@@

>>>>

\section{Complications Encountered While Writing The \parsernamelong{}}

@@

The fifteen complications documented in this section can be divided into
three broad classes:

\begin{description}

    \item [Recreating Postfix behaviour] Three complications are caused by
        \parsername{} accurately recreating Postfix's behaviour.

    \item [Lack of information in the log files] Seven complications are
        caused by lack of information in the log files.  In general some
        extra logging by Postfix could remove these complications, and in
        one case extra logging in later versions of Postfix does remove the
        complication (\sectionref{identifying bounce notifications}); three
        of the complications would be removed if Postfix generated a log
        line whenever it discarded a mail.

    \item [Order log lines are found in log files] Five complications are
        caused by out of order log lines.  Postfix is not to blame for
        these, they are caused by process scheduling, inter-process
        communication delays, and are more likely to occur when the mail
        server is heavily loaded.

@@

    \item Most importantly: each action needs to know whether it deals with
        accepted mails, and should find the mail in the by\_queueid state
        tables, or deals with delivery attempts, and thus will find the
        connection in the by\_pid state tables.  Some actions, e.g.\
        \action{DELIVERY\_REJECTED}, need to query both tables.  Support
        functions fall into two groups: those that deal with state tables,
        and so need to know which one to operate on; the other group are
        passed a specific mail or connection to work with, and do not need
        to know about state tables at all.

@@

Postfix does not log a message making the client's behaviour clear, so
again heuristics are required to identify when a delivery attempt is
aborted in this way.  A list of all mails accepted during a connection is
saved in the connection's state table entry, and the accepted mails are
examined when the disconnection action is invoked.  Each mail is checked
for the following:

@@

Some mail clients disconnect abruptly if a second or subsequent recipient
is rejected; they may also disconnect after other errors, but such
disconnections are either unimportant or are handled elsewhere in the
parser, e.g.\ \sectionref{timeouts during data phase}.  Postfix does not
log a message saying the mail has been discarded, as should be expected by
now.  The checks to identify this happening are:

@@

The author has only observed this complication occurring after a timeout,
though there may be other circumstances that trigger it.  Sometimes the
\daemon{cleanup} log line for a mail being accepted is logged after the
timeout log line, by which time \parsername{} has discarded the mail
because it will not have enough data to satisfy the database schema (see
\sectionref{timeouts during data phase}); parsing the \daemon{cleanup} log
line causes the \action{CLEANUP\_PROCESSING} action to create a new state
table entry, to help deal with re-injected mails (\sectionref{Re-injected
mails}).  This is incorrect because the log line actually belongs to the
mail that has just been discarded; if the queueid is reused there will be a
queueid clash, otherwise the new mail will just remain in the state tables.

@@

% \newpage{} % There's always a fucking varioref problem here

\subsection{Out Of Order Log Lines}

@@

\subsection{Mail Deleted Before Delivery Is Attempted}

@@

\subsection{Summary}

This section has described the complications encountered while implementing
\parsername{}.  Five of the fifteen complications are caused by log lines
appearing in an unexpected or abnormal order in log files, often because of
heavy system load: these complications typically caused warnings and a new
state table entry.  Three of the fifteen complications are required to
correctly recreate Postfix's behaviour.  The remaining seven complications
are caused by deficiencies in Postfix's logging, and some could easily be
resolved by adding additional logging to Postfix; these seven are
dealt with by applying heuristics to specific log lines and/or mails, in
some cases with the aid of information cached when mails are removed from
the state tables.

### 2009-04-06 doc/logparser-implementation.tex
Corrections/improvements based on Carl's comments.

    \item Reading and understanding the source code would require a
        significant investment of time: Postfix 2.5.5 has 17MB of source
        code.

@@

    \item This is followed by rejection of sender address, recipient
        addresses, client \acronym{IP} address or hostname, etc.\
        \flowchart{DELIVERY\_REJECTED}{4}; acceptance of one or more mails
        \flowchart{CLONE}{5}; or some random interleaving of both.

    \item The client disconnects \flowchart{DISCONNECT}{6}.  If Postfix has
        rejected any \acronym{SMTP} commands the resulting data will be
        saved to the database; if there were no rejections there will not
        be any data to save.  Any mails accepted will already have been
        cloned so their data is in another data structure, and will be
        delivered in the same way as mails that entered the system by any
        other route.

@@

    \item Postfix will log the sender and recipient addresses separately
        \flowchart{SAVE\_DATA}{8}.

@@

    \item Eventually the mail will be delivered, bounced, or deleted by the
        administrator \flowchart{COMMIT}{11}.  This is the last log line
        for this particular mail, though it may be indirectly referred to
        if it was re-injected.  The mail is cleaned up and entered in the
        database, then deleted from the state tables.

@@

    \item [restriction\_name] The name of the Postfix restriction that
        caused the mail to be rejected.  Only applicable to rules that
        recognise rejection log lines, i.e.\ rules that have an action of
        \action{DELIVERY\_REJECTED}; other rules should not set this field.

@@

    \item [enhanced\_status\_code] The enhanced status code~\cite{RFC3463}
        is similar to the \acronym{SMTP} code, but is intended to be
        interpreted by mail clients so that error messages can be clearly
        conveyed to the user.  Enhanced status code support was added to
        Postfix in version 2.3; log files from previous versions will not
        contain any enhanced status codes.

@@

The framework is now ready to begin parsing.  For each log line it will use
those rules whose \texttt{program} field equals the program in the log line
(see \sectionref{rule conditions in implementation}), falling back to
generic rules if necessary, and warning if the log line is unrecognised.
The rules are sorted for increased efficiency; see \sectionref{rule
ordering for efficiency} for details.  The repetitive nature of log files
gives them high compression ratios; the framework can read compressed log
files, simplifying the process of parsing them because they do not have to
be uncompressed to a temporary file before parsing begins.  When used
interactively, the framework also displays a progress bar to show how far
parsing has progressed through the log file and how long the remainder of
the parsing process is likely to take.  The progress bar is not as reliable
when parsing compressed log files, because a compressed block may
uncompress to a variable number of log lines; there is also variation
between the recognition and processing time of individual log lines, but
overall the progress bar is a useful addition.

@@

@@

    \item The regex from the recognising rule and the log line it
        recognised can be printed, so that the user can verify that the
        correct rule recognised each log line.  Individual rules can also
        set their debug field to true, and debugging information will be
        printed each time they recognise a log line.

@@

@@

@@

    \item [DELIVERY\_REJECTED] Deals with Postfix rejecting an
        \acronym{SMTP} command from the remote client.  This is the action
        most frequently specified by rules, because so many different
        restrictions are used to reject delivery attempts.  This action is
        quite simple: if the log line contains a queueid, save the data
        captured by the rule's regex to the mail identified by that
        queueid; otherwise save it to the connection identified by the
        \acronym{pid} in the log line.

@@

This chapter has presented \parsername{}, the parser implemented for this
project, beginning with the assumptions made during its development.  A
simplified flowchart shows the most common paths taken through Postfix and
\parsername{}, accompanied by a description of the stages and transitions.
A database provides storage for rules and for data gathered from log files;
any further use of that data is dependent on a clear understanding of the
database schema, so the resemblance between a database schema and an
\acronym{API} is described, followed by a detailed description and a
diagram of the database schema.  The framework is documented next,
including the steps it takes during initialisation, the parsing process,
and the conveniences it offers to users.  The performance data collected by
the framework is described, as are the ways in which various optimisations
can be disabled to demonstrate their effect, and the debugging options the
framework provides.  The implementation of actions in \parsername{} is
documented, including how frequently each action is specified by rules, and
brief descriptions of why some actions are more popular than others.  All
the actions that are part of \parsername{} are described in detail,
followed by the process of adding a new action to the parser.  The final
component of the architecture, the rules, is also the most visible
component, and its implementation is examined in detail with: an example
rule and an explanation of every field; a description of how to add new
rules and determine the values that should be used for each field; an
explanation of the algorithm used by the utility that creates new regexes
based on unparsed log lines, and how it differs from the original algorithm
it is based on.  How \parsername{} uses rule conditions and overlapping
rules is discussed, accompanied by a description of the regex snippets the
framework provides to simplify the regexes used in rules.  The many
complications and difficulties encountered while writing \parsername{}, and
the solutions developed to overcome them, are documented in depth; also
documented are how some solutions interact, and which action or actions the
solution is implemented in.  This chapter concludes with a list of the
limitations identified in \parsername{}, followed by some possible
improvements suggested by the author.

>>>>

    \item Reading and understanding the source code would require a
        significant investment of time: Postfix 2.5.5 has 17MB of source
        code.  Each subsequent version would require further work to
        investigate the changes; many of those changes, although they
        improve Postfix's internals, would not have any effect on its
        externally observable behaviour.

@@

    \item This is followed by rejection of sender address, recipient
        addresses, client \acronym{IP} address or hostname, etc.\
        \flowchart{DELIVERY\_REJECTED}{4}; acceptance of one or more mails
        \flowchart{CLONE}{5}; failure of the remote client's connection
        \flowchart{DELIVERY\_ERROR}{6}; or some random interleaving of two
        or more of the above.

    \item The client disconnects \flowchart{DISCONNECT, TIMEOUT}{6}, either
        normally or with an error.  If Postfix has rejected any
        \acronym{SMTP} commands the resulting data will be saved to the
        database; if there were no rejections there will not be any data to
        save.  Any mails accepted will already have been cloned so their
        data is in another data structure, and will be delivered in the
        same way as mails that entered the system by any other route.

@@

    \item The mail will have one or more delivery attempts
        \flowchart{MAIL\_SENT, SAVE\_DATA}{8}.

@@

    \item After one or more delivery attempts the mail will be delivered
        \flowchart{MAIL\_SENT}{8}, bounced \flowchart{MAIL\_BOUNCED}{8},
        expired \flowchart{EXPIRY}{8}, or deleted by the administrator
        \flowchart{DELETE}{8}.

    \item This is the last log line for this particular mail, though it may
        be indirectly referred to if it was re-injected.  The mail is
        cleaned up and entered in the database, then deleted from the state
        tables \flowchart{COMMIT}{11}.

@@

    \item [restriction\_name] The name of the Postfix restriction that
        caused the mail to be rejected.  This field is meaningful only for
        rules that recognise rejection log lines, i.e.\ rules that have an
        action of \action{DELIVERY\_REJECTED}.

@@

    \item [enhanced\_status\_code] The enhanced status code~\cite{RFC3463}
        is similar to the \acronym{SMTP} code, but is intended to be
        interpreted by mail clients so that error messages can be clearly
        conveyed to the user.  Enhanced status code support was added to
        Postfix in version 2.3; log lines from previous versions will not
        contain any enhanced status codes.

@@

The framework is now ready to begin parsing.  For each log line it will use
those rules whose \texttt{program} field equals the program in the log line
(previously described in \sectionref{rule conditions in implementation}),
falling back to generic rules if necessary, and warning if the log line is
unrecognised.  The rules are sorted for increased efficiency; see
\sectionref{rule ordering for efficiency} for details.  The repetitive
nature of log files gives them high compression ratios; the framework uses
a Perl module named \texttt{IO::Uncompress::AnyUncompress} to read
compressed log files; this saves users the trouble of uncompressing them to
a temporary file before parsing begins.  When used interactively, the
framework also displays a progress bar to show how far parsing has
progressed through the log file and how long the remainder of the parsing
process is likely to take.  The progress bar is not as reliable when
parsing compressed log files, because a compressed block may uncompress to
a variable number of log lines; there is also variation between the
recognition and processing time of individual log lines, but overall the
progress bar is a useful addition.

@@

XXX CARL SUGGESTS ADDING GRAPHS AND STATS TABLES FOR THESE\@.

@@

    \item Individual rules can set their debug field to true, and debugging
        information will be printed each time they recognise a log line.

@@

    \item Every time a log line is recognised, the recognising rule's regex
        and the log line can be printed, so that the user can verify that
        the correct rule recognised each log line.

@@

A user, typically a mail administrator, would use these options when having
difficulty adding or modifying rules to recognise log lines not recognised
by \parsernames{} \numberOFrules{} rules.  Option 1 is useful for debugging
a single rule, to check if it is recognising log lines it should not;
figuring out that a rule is not recognising log lines that it should
recognise is a more difficult task.  Option 2 is useful when \parsername{}
is warning about parsing problems, because all the information about a mail
will be present in the warning message: the full text of every log line
that is part of the mail; the rule that recognised each log line; the
filename of the log file and the line number within it for each log line.
Option 3 is less useful because of the volume of data it produces, though
for small log file snippets, e.g.\ 1000 log lines, it is possible to
manually verify that the correct rule recognised each log line.  Option 4
is not useful for most users: it is for debugging problems with the data
that is saved to the database, and there are sufficient safeguards in place
that there should be plenty of warning messages explaining what is wrong
without using this option.

@@

    \item [DELIVERY\_REJECTED] Postfix rejected an \acronym{SMTP} command
        from the remote client.  This is the action most frequently
        specified by rules, because so many different restrictions are used
        to reject delivery attempts.  This action is quite simple: if the
        log line contains a queueid, save the data captured by the rule's
        regex to the mail identified by that queueid; otherwise save it to
        the connection identified by the \acronym{pid} in the log line.

@@

This chapter has presented \parsername{}, the parser implemented for this
project, beginning with the assumptions made during its development.  A
simplified flowchart shows the most common paths taken through Postfix and
\parsername{}, accompanied by a description of the stages and transitions.
A database provides storage for rules and for data gathered from log files;
any further use of that data is dependent on a clear understanding of the
database schema, so the resemblance between a database schema and an
\acronym{API} is described, followed by a detailed description and a
diagram of the database schema.  The framework is documented next,
including the steps it takes during initialisation, the parsing process,
and the conveniences it offers to users.  The performance data collected by
the framework is described, as are the ways in which various optimisations
can be disabled to demonstrate their effect, and the debugging options the
framework provides.  The implementation of actions in \parsername{} is
documented, including how frequently each action is specified by rules, and
brief descriptions of why some actions are more popular than others.  All
the actions that are part of \parsername{} are described in detail,
followed by the process of adding a new action to the parser.  The final
component of the architecture, the rules, is also the most visible
component, and its implementation is examined in detail with: an example
rule and an explanation of every field; a description of how to add new
rules and determine the values that should be used for each field; an
explanation of the algorithm used by the utility that creates new regexes
based on unparsed log lines, and how it differs from the original algorithm
it is based on.  How \parsername{} uses rule conditions and overlapping
rules is discussed, accompanied by a description of the regex snippets the
framework provides to simplify the regexes used in rules.  The many
complications and difficulties encountered while writing \parsername{}, and
the solutions developed to overcome them, are documented in depth; also
documented are how some solutions interact, and which action or actions the
solution is implemented in.  This chapter concludes with a list of the
limitations identified in \parsername{}, followed by some possible
improvements suggested by the author.  The next chapter will evaluate this
implementation, examining both \parsernames{} efficiency and the coverage
it achieves when parsing log files.
